---
title: "ST558 Project 2"
author: "Alex Selwaeh and Zichang Xiang"
date: "7/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction Section
The purpose of this project is to create predictive models and automate R Markdown reports. The data we will use is the number of Bike-sharing users aggregated on daily basis from the Capital Bikeshare system in 2011 and 2012. In this data set, there are 16 variables that are related to bike rental counts. The variables we will use include season, year (yr), month (mnth), hour (hr), holiday, weekday, workingday, weathersit, windspeed, casual, registered, and count (cnt). Since some of our predictors are categorical and our response is count, we will use the generalized linear model to predict the response.

# Data
```{r message=FALSE}
#load packages
library(readr)
library(dplyr)
```


```{r message=FALSE}
#read in data
day <- read_csv("day.csv")
day
#data for Monday
Mon <- day %>% filter(weekday == 1)
Mon
```


```{r}
#split data set into training and test sets
set.seed(1)
train <- sample(1:nrow(Mon), size = nrow(Mon)*0.7)
test <- dplyr::setdiff(1:nrow(Mon), train)

train <- Mon[train, ]
test <- Mon[test, ]
train
test
```

# Summarizations
```{r}

```

# Modeling
```{r}

```

