---
title: "ST558 Project 2"
author: "Alex Selwaeh and Zichang Xiang"
date: "7/1/2021"
output: html_document  
params: 
      dow: "Monday"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction Section

The purpose of this project is to create predictive models and automate R Markdown reports. The data we will use is the number of Bike-sharing users aggregated on daily basis from the Capital Bikeshare system in 2011 and 2012. In this data set, there are 16 variables that are related to bike rental counts. The variables we will use include season, year (yr), mnth, holiday, weekday, workingday, weathersit, hum, atemp, temp, casual, registered, and count (cnt). To model response counts (cnt), we will use linear regression models and ensemble tree methods (random forest method and boosted tree method).

## Data
```{r message=FALSE}
#load packages
library(corrplot)
library(ggplot2)
library(cowplot)
library(modelr)
library(readr)
library(dplyr)
library(knitr)
library(caret)
library(tidyr)
library(purrr)
library(gbm)

```


```{r message=FALSE}
#read in data
dayData <- read_csv("day.csv")
dayData

#check for missing values
anyNA(dayData)

#create for loop to subset data for each weekday
status<-vector()

for (i in seq_len(nrow(dayData))){
  if(dayData$weekday[i] == 1){
    status[i] <- "Monday"
  } else if (dayData$weekday[i] == 2){
    status[i] <- "Tuesday"
  } else if (dayData$weekday[i] == 3){
    status[i] <- "Wednesday"
  } else if (dayData$weekday[i] == 4){
    status[i] <- "Thursday"
  } else if (dayData$weekday[i] == 5){
    status[i] <- "Friday"
  } else if (dayData$weekday[i] == 6){
    status[i] <- "Saturday"
  } else {
    status[i] <- "Sunday"
  }
}

dayData$status <- status
#data for Monday
paramsData <- dayData %>% filter(status == params$dow)

#Create columns to represent the categorical columns as mentioned in READ.ME
paramsData <- paramsData %>%
  # Add columns to represent the categorical columns as mentioned in READ.ME.
  mutate(SeasonType = ifelse(season == 1, "spring", ifelse(season == 2, "summer", ifelse(season == 3, "fall", "winter"))), yearType= ifelse(yr == 0, "2011", "2012"), workingdayType= ifelse(workingday == 1, "Working Day", "Non WorkingDay"),weathersitType= ifelse(weathersit == 1, "Clear", ifelse(weathersit == 2, "Mist", ifelse(weathersit == 3, "Light Snow", "HeavyRain"))))

#convert month from numerical to categorical charcter  
paramsData$mnth1 <- as.character(paramsData$mnth)

```


```{r}
#split data set into training and test sets
set.seed(1)
train <- sample(1:nrow(paramsData), size = nrow(paramsData)*0.7)
test <- dplyr::setdiff(1:nrow(paramsData), train)

train <- paramsData[train, ]
test <- paramsData[test, ]

train
test
```


## Summarizations

### Summary Statistics

Summary statistics give us a quick look of our data. For our case, we can find out the average number of bike rentals per season.

```{r}
# Create a table of summary stats.
seasonSummary <- train %>% 
  # Select the seasone and cnt columns.
  select(SeasonType, cnt) %>%
  # Group by season
  group_by(SeasonType) %>%
  # Get summary statistics for total users by season.
  summarize("Min." = min(cnt),
            "1st Quartile" = quantile(cnt, 0.25),
            "Median" = quantile(cnt, 0.5),
            "Mean" = mean(cnt),
            "3rd Quartile" = quantile(cnt, 0.75),
            "Max" = max(cnt),
            "Std. Dev." = sd(cnt)
            )

# Display a table of the summary stats.
knitr::kable(seasonSummary, 
             caption=paste("Summary Statistics for total users",
                           "By Season"),
             digits=2)
```

#### Contingency Tables

A continuity table shows the relationship between two categorical variables. In our case, we can determine whether the season and weather are related, and whether the season and workday are related.

```{r}
#create contingency tables
kable(table(train$SeasonType, train$weathersitType))
kable(table(train$SeasonType, train$workingdayType))
```

### Plots
#### Correlation Plot

Correlation plot shows the strength of a relationship between two variables. In our case, we can identify which variables are highly correlated with one another, especially with the response, the number of bikes rented. 

```{r}
#create correlation plot
corr <- cor(train[, -c(1,2,7,17:22)])
head(round(corr, 2))
corrplot(corr, type = "upper", method = "pie")

```


#### Histograms

Histograms are used to summarize distributions of variables. In our case, we are trying to find out whether the change in each variable has an impact on the number of bikes rented.

```{r message=FALSE}
#create boxplots for selected variables (referenced from https://drsimonj.svbtle.com/quick-plot-of-all-variables)
#reshape the data set
reshape <- train %>% keep(is.numeric) %>% gather() 
#plot the density plot
g <- ggplot(reshape, aes(x = value))
g + facet_wrap(~ key, scales = "free") + geom_density()

```


#### Boxplots

Boxplots show the shape of the distribution of each variable. By looking at the boxplots below, we can see how each variable affects the number of bikes rented. 


```{r message=FALSE}
#create boxplot for each variables
attach(train)
par(mfrow = c(1,4))
boxplot(season, xlab = "season")
boxplot(mnth, xlab =" mnth")
boxplot(weathersit, xlab = "weathersit")
boxplot(windspeed, xlab = " windspreed")
boxplot(yr, xlab = "yr")
boxplot(temp, xlab = "temp")
boxplot(atemp, xlab = "atemp")

#remove outliers
train <- train[!abs(train$windspeed)>0.4, ]
```


Boxplot with the number of users on the y-axis (wether casual,registered or total users) and the season on the x-axis - We can inspect the trend of users across seasons using these plots. Notice that the biggest contribution towards total number of users comes from the registered users which is expected. The most active seasons for that `r params$dow` is the fall season and the least active season is the spring.  


```{r}
#create boxplots
plot11 <- ggplot(train, aes(SeasonType,
                                   cnt,
                                   color=cnt)) +
  geom_boxplot() + 
  # Jitter the points to add a little more info to the boxplot.
  geom_jitter() + 
  # Add labels to the axes.
  scale_x_discrete("Season") + scale_y_continuous("Total Users") +
  ggtitle("Total Users by Season") + theme(legend.position="none")
plot11
plot12 <- ggplot(train, aes(SeasonType,
                                   casual,
                                   color=cnt)) +
  geom_boxplot() + 
  # Jitter the points to add a little more info to the boxplot.
  geom_jitter() + 
  # Add labels to the axes.
  scale_x_discrete("Season") + scale_y_continuous("Casual Users") +
  ggtitle("Casual Users by Season") + theme(legend.position="none")
plot12
plot13 <- ggplot(train, aes(SeasonType,
                                   registered,
                                   color=cnt)) +
  geom_boxplot() + 
  # Jitter the points to add a little more info to the boxplot.
  geom_jitter() + 
  # Add labels to the axes.
  scale_x_discrete("Season") + scale_y_continuous("Registered Users") +
  ggtitle("Registered Users by Season") + theme(legend.position="none")
plot13
plot_grid(plot13, plot12,plot11, ncol=3)

```


#### Scatter plots

Scatter plot with the total number of users on the y-axis and the temperature, wind speed, humidity on the x-axis - We can inspect the trend of total users across these variables and notice that humidity almost has a negligible effect on total number of users. Also it's noticeable that the wind speed has a negative effect on the total number of users where as the wind speed increases, the number of users will decrease. 

```{r}
#create scatter plots
plot21 <- ggplot(train, aes(temp,
                                   cnt,
                                   color=cnt)) + 
  geom_point(size=4, alpha=0.75) + scale_color_gradient(low="blue", high="red") + 
  theme(legend.position="none") + geom_smooth(method=lm, formula=y~x, color="black") + 
  scale_x_continuous("Temprature") + 
  scale_y_continuous("Total Users") + 
  ggtitle("Temprature. vs. Total Users")
plot21

plot22 <- ggplot(train, aes(windspeed,
                                   cnt,
                                   color=cnt)) + 
  geom_point(size=4, alpha=0.75) + scale_color_gradient(low="blue", high="red") + 
  theme(legend.position="none") + geom_smooth(method=lm, formula=y~x, color="black") + 
  scale_x_continuous("Wind Speed") + 
  scale_y_continuous("Total Users") + 
  ggtitle("Wind Speed vs. Total Users")
plot22

plot23 <- ggplot(train, aes(hum,
                                   cnt,
                                   color=cnt)) + 
  geom_point(size=4, alpha=0.75) + scale_color_gradient(low="blue", high="red") + 
  theme(legend.position="none") + geom_smooth(method=lm, formula=y~x, color="black") + 
  scale_x_continuous("Humidity") + 
  scale_y_continuous("Total Users") + 
  ggtitle("Humidity vs. Total Users")
plot23

plot_grid(plot21, plot22, plot23,ncol=3)
```


scatterplot with the number of users on the y-axis and the month on the x-axis, We
can inspect the trend of users across months using this plot.

```{r}
plot31 <- ggplot(paramsData, aes(x = mnth, y = casual)) + geom_point() +
    geom_smooth() +
    geom_smooth(method = lm, col = "Red")
plot31
plot32 <- ggplot(paramsData, aes(x = mnth, y = registered)) + geom_point() +
    geom_smooth() +
    geom_smooth(method = lm, col = "Red")
plot32
plot33 <- ggplot(paramsData, aes(x = mnth, y = cnt)) + geom_point() +
    geom_smooth() +
    geom_smooth(method = lm, col = "Red")
plot33
plot_grid(plot31, plot32, plot33,ncol=3)
```


## Modeling
### Modeling of the first group member
```{r}
# select models
fit1 <- lm(cnt ~ windspeed + hum + atemp + temp + weathersit + workingday + holiday + mnth + yr + season, data = train)
fit1
# use anova table to choose the significant variables and reduce the model selection
anova(fit1)
# anova reveals that season,yr,temp and atemp has a p-value of significance, therefore we will choose only these 4 variables in our analysis.
summary(fit1)
# start reducing fit1 by choosing the most significant variables based on p-value less than 10% considered highly significant
fit2 <- lm(cnt ~ season+yr+temp+atemp+windspeed+mnth , data = train)
fit2
anova(fit2)
summary(fit2)
# Since in fit2 the anova revealed that atemp is not as highly significant as the rest of the variables in fit2, therefore reduce more
#also note that temp and atemp are highly correlated predictors causing multicollinearity therefore it's safe to drop atemp
fit3 <- lm(cnt ~ season+yr+temp+windspeed+mnth , data = train)
fit3
anova(fit3)
# the data shows that we are interested in the following variables
#temp,yr,season, windspeed,mnth and cnt as the response
#Explore the full chosen data correlations
corrData <- train %>% select(cnt,season,yr,temp,windspeed,mnth)
GGally::ggpairs(corrData)
#Correlation data table shows that wind speed and nth have the lowest correlation when compared with temperature, year and season. Therefore, we will reduce our variables to temp,yr,season.
#Now start exploring among these 5 variables for the best fit model
glm1Fit <- lm(cnt ~ season+yr+temp, data = train)
glm2Fit <- lm(cnt ~ temp:yr, data = train)
glm3Fit <- lm(cnt ~ temp:season, data = train)
glm4Fit <- lm(cnt ~ temp*yr*season, data = train)
glm5Fit <- lm(cnt ~ temp+yr+I(yr^2), data = train)

model <- c(("glm1Fit"),("glm2Fit"),("glm3Fit"),("glm4Fit"),("glm5Fit"))
  
trainMSE <- c(rmse(glm1Fit, train),
rmse(glm2Fit, train),
rmse(glm3Fit, train),
rmse(glm4Fit, train),
rmse(glm4Fit, train)
)
testMSE <- c(rmse(glm1Fit, test),
rmse(glm2Fit, test),
rmse(glm3Fit, test),
rmse(glm4Fit, test),
rmse(glm5Fit, test)
)
MSEdf <- data.frame(model, trainMSE, testMSE)
MSEdf <- MSEdf %>% arrange(testMSE)
MSEdf

# therefore, this is the chosen model glm4Fit
#glm4Fit <- lm(cnt ~ temp*yr*season, data = train)
#might explore more on top 3 fits which are glm4Fit,glm1it,glm5Fit
```

### Modeling of the second group member

```{r}
#use step function to select variables
select <- step(lm(cnt ~ 1, train), scope = cnt ~ season + yr + mnth + temp + atemp, direction = "both", k = log(nrow(train))) # here we use BIC because we set k equals to log(n)
#view the selected variables
colnames(select$model)
```

Considering the above result, the selected variables are cnt, atemp, yr, and season, as the model that includes these variables has the smallest BIC.

```{r}
#select models
set.seed(1)
fit1 <- train(cnt ~ season + yr  + workingday + weathersit + temp + atemp + windspeed, 
              data = train,
              method = "lm",
              preProcess = c("center", "scale"),
              trControl = trainControl(method = "cv", number = 10)
              )

fit2 <- train(cnt ~ season + yr + atemp, 
              data = train,
              method = "lm",
              preProcess = c("center", "scale"),
              trControl = trainControl(method = "cv", number = 10)
                )

fit3 <- train(cnt ~ season*yr*atemp, 
              data = train,
              method = "lm",
              preProcess = c("center", "scale"),
              trControl = trainControl(method = "cv", number = 10)
                )

fit4 <- train(cnt ~ I(season)^2 + I(yr)^2 + I(atemp)^2,
              data = train,
              method = "lm",
              preProcess = c("center", "scale"),
              trControl = trainControl(method = "cv", number = 10)
              )

fit5 <- train(cnt ~ I(yr)^2 + season *atemp,
              data = train,
              method = "lm",
              preProcess = c("center", "scale"),
              trControl = trainControl(method = "cv", number = 10)
              )

#compare each model
kable(data.frame(t(fit1$results), t(fit2$results), t(fit3$results), t(fit4$results), t(fit5$results)))

#check the fit
lmPred1 <- predict(fit1, newdata = test)
lmFit1 <- postResample(lmPred1, obs = test$cnt)

lmPred2 <- predict(fit2, newdata = test)
lmFit2 <- postResample(lmPred2, obs = test$cnt)

lmPred3 <- predict(fit3, newdata = test)
lmFit3 <- postResample(lmPred3, obs = test$cnt)

lmPred4 <- predict(fit4, newdata = test)
lmFit4 <- postResample(lmPred4, obs = test$cnt)

lmPred5 <- predict(fit5, newdata = test)
lmFit5 <- postResample(lmPred5, obs = test$cnt)
lmRMSE <- kable(data.frame(lmFit1[1], lmFit2[1], lmFit3[1], lmFit4[1], lmFit5[1]))
lmRMSE
#final model: fit5: cnt ~ I(yr)^2 + season *atemp
```

Five candidate linear regression models were fitted using the `train()` function and 10-fold cross-validation. In comparing their RMSE on the test sets, I chose the final model fit5 (cnt ~ I(yr)^2 + season *atemp), which has the lowest RMSE. 

We use the train() function to fit the boosted tree model. First, we must give the model and data set we use. Next, we provide the method we use, which is “gbm”. Then, we use the `preProcess()` function to standardize the data. Finally, we specify the type of cross-validation we wish to perform. In our case, we would like to perform repeated cross-validation with 10 folds for 5 times.

```{r results='hide'}
#fitting boosted tree model for counts
set.seed(1)
boostFit <- train(cnt ~ I(yr)^2 + season *atemp, 
                  data = train, 
                  method = "gbm", 
                  preProcess = c("center", "scale"),
                  trControl = trainControl(method = "repeatedcv", number = 10, repeats = 5))
```

```{r}
#view the results
boostFit$results
#view the best model
boostFit$bestTune
#predict cnt and calculate RMSE
boostPred <- predict(boostFit, newdata = test)
result <- postResample(boostPred, test$cnt)
boostRMSE <- result[1]
boostRMSE
```

### Comparison of all four models
```{r}
results <- data.frame(MSEdf[1,3], lmFit5[1], boostRMSE)
colnames(results) <- c("lmFit_1stMember", "lmFit_2ndMember", "Boost Tree")
kable(results)
```